---
title: "CS-ARDL in R: Estimating Long-Run Effects with Cross-Sectional Dependence"
author: "Joao Claudio Macosso"
description: >
  A technical guide to specifying, estimating, diagnosing, and reporting
  cross-sectionally augmented ARDL models in R, with a published replication.
bibliography: "`r system.file('REFERENCES.bib', package = 'csdm')`"
csl: "`r system.file('apa.csl', package = 'csdm')`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE,
  fig.align = "center"
)
options(width = 90)
library(csdm)

helper <- file.path("vignettes", "articles", "cmpr-data.R")
if (!file.exists(helper)) helper <- "cmpr-data.R"
source(helper, local = knitr::knit_global())
```

## The applied problem

Suppose a thesis asks whether inflation and persistent increases in public debt
affect output growth in the long run. A pooled dynamic panel would require the
same slopes for every country and would leave global shocks untreated. Separate
country regressions allow heterogeneous slopes, but their errors can remain
correlated through recessions, commodity prices, financial conditions, or other
unobserved common factors.

The cross-sectionally augmented autoregressive distributed lag estimator
(CS-ARDL) addresses both problems. It estimates a dynamic regression for each
unit, augments it with cross-sectional averages, transforms the estimated ARDL
coefficients into unit-specific long-run effects, and then averages those
effects across units. This guide develops that calculation and reproduces Table
17 of @ChudikEtAl2013. The method was subsequently published and developed in
@ChudikEtAl2016. Ditzen's `xtdcce2` provides an independent Stata implementation
[@Ditzen2018].

## Model and estimand

For unit $i$, let $y_{it}$ be the dependent variable, let $\mathbf{x}_{it}$
contain $K$ regressors, and let $\bar{\mathbf z}_t$ collect cross-sectional
averages. A CS-ARDL$(p,q)$ regression is

$$
y_{it}=c_i+
\sum_{\ell=1}^{p}\phi_{i\ell}y_{i,t-\ell}+
\sum_{\ell=0}^{q}\boldsymbol\beta_{i\ell}'\mathbf{x}_{i,t-\ell}+
\sum_{\ell=0}^{p_{\bar z}}\boldsymbol\delta_{i\ell}'
\bar{\mathbf z}_{t-\ell}+e_{it}.
$$

Define the autoregressive denominator

$$
d_i=1-\sum_{\ell=1}^{p}\phi_{i\ell}.
$$

The unit-specific adjustment coefficient and long-run effects are

$$
\lambda_i=-d_i,
\qquad
\boldsymbol\theta_i=
\frac{\sum_{\ell=0}^{q}\boldsymbol\beta_{i\ell}}{d_i}.
$$

`csdm` calculates these nonlinear ratios for each eligible unit and then forms
the mean-group estimates

$$
\widehat{\boldsymbol\theta}_{MG}
=\frac{1}{N_e}\sum_{i\in\mathcal E}\widehat{\boldsymbol\theta}_i,
$$

where $\mathcal E$ is the common set of units for which all requested
parameters are finite. The mean-group covariance is the cross-unit sample
covariance of the unit estimates divided by $N_e$. This is large-$N$ inference;
it does not remove the finite-$T$ bias in the unit dynamic regressions.

Two lag choices have different jobs:

- `csdm_lr(ylags = p, xdlags = q)` sets the ARDL orders.
- `csdm_csa(lags = p_z)` sets the maximum lag of the cross-sectional averages.

In the replication, the ARDL orders are $(1,1)$, $(2,2)$, and $(3,3)$, while
the cross-sectional-average lag order remains three in every model.

## Data, provenance, and licensing

The application uses annual GDP, CPI, and gross government debt-to-GDP data for
40 countries from 1965 through 2010. The files are deposited in the
[UK Data Service ReShare record](https://reshare.ukdataservice.ac.uk/851454/)
(DOI: [10.5255/UKDA-SN-851454](https://doi.org/10.5255/UKDA-SN-851454)).

The deposit is licensed under
[CC BY-NC-SA 4.0](https://creativecommons.org/licenses/by-nc-sa/4.0/). It is not
distributed in `csdm` or in the package repository. During the website build,
the article downloads the archive from the institutional record, verifies its
SHA-256 checksum, and reads it from a temporary cache. The package code remains
GPL-3. Attribution and the source-data license continue to apply to the source
files and to derived material where required.

For an offline build, point `CSDM_CMPR_PATH` to `CMPR.dta`, the downloaded ZIP,
or the directory containing `CMPR.dta`:

```r
Sys.setenv(CSDM_CMPR_PATH = "/path/to/CMPR_Data")
```

```{r acquire-data}
cmpr <- prepare_cmpr()

data.frame(
  countries = length(unique(cmpr$ccode)),
  first_year = min(cmpr$year),
  last_year = max(cmpr$year),
  rows = nrow(cmpr),
  archive_sha256 = CMPR_ARCHIVE_SHA256
)
```

The transformations follow the supplied Stata do-file:

$$
y_{it}=\log(GDP_{it}),\quad
\Delta y_{it}=y_{it}-y_{i,t-1},
$$

$$
\pi_{it}=\Delta\log(CPI_{it}),\quad
\Delta d_{it}=\Delta\log(Debt/GDP_{it}).
$$

The estimated dependent variable is `dy`, output growth. The regressors are
`dp`, inflation, and `dgd`, the growth rate of the debt-to-GDP ratio. Panel
differences are calculated only across consecutive years and are never allowed
to cross country boundaries.

```{r transformed-data}
summary(cmpr[c("dy", "dp", "dgd")])
```

## Specifying the nine CS-ARDL models

The paper considers three economic specifications at each lag order:

- **(a)** debt-to-GDP growth only;
- **(b)** inflation only;
- **(c)** debt-to-GDP growth and inflation jointly.

The authors calculate each cross-sectional average from every available value
of that variable before ARDL lag trimming. Because the variables have different
missing-value patterns, `fullsample = TRUE` is necessary for this replication.

```{r fit-function}
fit_cmpr_cs_ardl <- function(regressors, order) {
  csdm(
    formula = reformulate(regressors, response = "dy"),
    data = cmpr,
    id = "ccode",
    time = "year",
    model = "cs_ardl",
    csa = csdm_csa(vars = c("dy", regressors), lags = 3),
    lr = csdm_lr(type = "ardl", ylags = order, xdlags = order),
    fullsample = TRUE
  )
}
```

```{r fit-models}
economic_specifications <- list(
  a = "dgd",
  b = "dp",
  c = c("dgd", "dp")
)

fits <- list()
for (order in 1:3) {
  for (specification in names(economic_specifications)) {
    name <- paste0(specification, order)
    fits[[name]] <- fit_cmpr_cs_ardl(
      economic_specifications[[specification]],
      order
    )
  }
}
```

A common mistake is to set `xdlags = 0` for these models. Table 17 uses the
same order for the dependent variable and each regressor. For example, its
one-lag model is an ARDL$(1,1)$, so both `ylags` and `xdlags` equal one.

## Reproducing Table 17

`csdm_fit` objects implement `tidy()` and `glance()`, so `modelsummary` can
consume the fitted models directly. The table keeps the coefficient names used
by the models: `lr_dgd` and `lr_dp` are the long-run coefficients, while
`lr_dy` is the adjustment coefficient. The remaining level and lag
coefficients are available in the fitted objects but omitted here to match the
focus of Table 17.

```{r modelsummary-table, results='asis'}
modelsummary::modelsummary(
  fits,
  coef_map = c(
    lr_dgd = "lr_dgd",
    lr_dp = "lr_dp",
    lr_dy = "lr_dy"
  ),
  statistic = "({std.error})",
  stars = TRUE,
  fmt = modelsummary::fmt_decimal(digits = 3, pdigits = 3),
  gof_map = c("nobs", "cd.statistic"),
  output = "markdown"
)
```

All nine models reproduce the published coefficients, standard errors,
adjustment estimates, CD statistics, and observation counts at the reported
precision. Matching the observation counts is especially useful: it verifies
the transformations, lag orders, cross-sectional-average scope, and missing-data
handling together.

The following hidden check compares the fitted results with the values printed
in the paper. It stops the website build if a result falls outside the relevant
rounding interval.

```{r extract-results, include=FALSE}
extract_replication_row <- function(fit, model) {
  long_run <- broom::tidy(fit, component = "long_run")
  adjustment <- broom::tidy(fit, component = "adjustment")
  cd <- cd_test(fit, type = "CD")$tests$CD

  value <- function(term, column) {
    found <- long_run[[column]][long_run$term == term]
    if (length(found)) found else NA_real_
  }

  data.frame(
    model = model,
    debt = value("lr_dgd", "estimate"),
    debt_se = value("lr_dgd", "std.error"),
    inflation = value("lr_dp", "estimate"),
    inflation_se = value("lr_dp", "std.error"),
    adjustment = adjustment$estimate,
    adjustment_se = adjustment$std.error,
    cd = cd$statistic,
    observations = nobs(fit),
    row.names = NULL
  )
}

estimated <- do.call(
  rbind,
  Map(extract_replication_row, fits, names(fits))
)
```

```{r verify-published-results, include=FALSE}
published <- data.frame(
  model = c("a1", "b1", "c1", "a2", "b2", "c2", "a3", "b3", "c3"),
  debt = c(-0.087, NA, -0.087, -0.090, NA, -0.079, -0.096, NA, -0.120),
  debt_se = c(0.013, NA, 0.016, 0.013, NA, 0.022, 0.016, NA, 0.040),
  inflation = c(NA, -0.083, -0.164, NA, -0.071, -0.110, NA, -0.065, -0.080),
  inflation_se = c(NA, 0.034, 0.038, NA, 0.031, 0.035, NA, 0.041, 0.059),
  adjustment = c(-0.889, -0.790, -0.952, -0.967, -0.817, -1.058, -0.920, -0.792, -1.210),
  adjustment_se = c(0.031, 0.041, 0.039, 0.042, 0.053, 0.053, 0.047, 0.058, 0.201),
  cd = c(-0.94, -0.30, 0.55, -0.43, 0.02, -0.11, -0.21, 0.05, -0.56),
  observations = c(1599, 1657, 1599, 1581, 1652, 1581, 1562, 1645, 1562)
)

estimated <- estimated[match(published$model, estimated$model), ]
inside_rounding_interval <- function(actual, printed, digits) {
  both_missing <- is.na(actual) & is.na(printed)
  close <- abs(actual - printed) < 0.5 * 10^(-digits) + 1e-12
  both_missing | (!is.na(close) & close)
}

three_digit_columns <- setdiff(names(published), c("model", "cd", "observations"))
checks <- vapply(
  three_digit_columns,
  function(column) inside_rounding_interval(
    estimated[[column]], published[[column]], 3
  ),
  logical(nrow(published))
)
stopifnot(
  all(checks),
  all(inside_rounding_interval(estimated$cd, published$cd, 2)),
  all(estimated$observations == published$observations)
)
```

## Lag sensitivity and diagnostics

Lag-order robustness should be evaluated on the estimates, uncertainty, sample,
and dynamics together. For joint specification (c), increasing the ARDL order
uses more initial observations for lags and therefore reduces the estimation
sample. The three-lag model also produces a noticeably less precise adjustment
estimate and a larger debt effect in absolute value.

```{r stability-table}
joint <- estimated[substr(estimated$model, 1, 1) == "c", ]
joint$stable_units <- vapply(
  fits[c("c1", "c2", "c3")],
  function(fit) sum(fit$cs_ardl$stable, na.rm = TRUE),
  integer(1)
)
joint$eligible_units <- vapply(
  fits[c("c1", "c2", "c3")],
  function(fit) fit$components$long_run$n_used,
  integer(1)
)

lag_display <- data.frame(
  Model = names(fits[c("c1", "c2", "c3")]),
  `Stable/eligible units` = paste0(
    joint$stable_units, "/", joint$eligible_units
  ),
  check.names = FALSE
)

knitr::kable(
  lag_display,
  align = c("c", "r"),
  row.names = FALSE,
  caption = "Lag sensitivity for the joint debt-inflation specification."
)
```

All 40 units contribute finite long-run estimates in each model. In the
three-lag specification, however, three unit AR polynomials are flagged as
unstable. `csdm` reports that diagnostic and does not silently discard those
units, so the estimate should be presented together with its stability result
and the lower-order specifications.

The paper reports the classical Pesaran CD statistic. For example:

```{r residual-cd}
cd_test(fits$c1, type = "CD")
```

The large positive CD statistics from unaugmented ARDL models in the paper fall
close to zero after cross-sectional augmentation. A small CD statistic is useful
evidence about remaining average pairwise residual correlation, but it does not
prove that every form of strong or weak dependence has been removed.

## Interpretation and limits

The replicated long-run estimates are negative. In specification (c), the
estimated debt-growth effect ranges from approximately -0.079 to -0.120, and the
inflation effect ranges from approximately -0.080 to -0.164. These are averages
of country-specific nonlinear long-run ratios, not coefficients from one pooled
regression.

Substantive interpretation requires several checks:

1. **Dynamic stability.** A long-run ratio is meaningful only when its AR
   denominator is nonzero and the fitted dynamics support a stable long-run
   interpretation. Inspect `fit$cs_ardl$stable` and the stored AR roots.
2. **Lag sensitivity.** Long-run ARDL ratios can change materially with the lag
   order. Report the sensitivity rather than selecting one order only because it
   produces a preferred significance level.
3. **Finite-$T$ bias.** Cross-sectional augmentation addresses common factors;
   it does not eliminate bias in short dynamic unit regressions. The adjustment
   estimates should therefore be interpreted cautiously.
4. **Factor coverage.** Cross-sectional averages must span the relevant common
   factor space. Three lags reproduce this application, but they are not a
   universal default.
5. **Long-run claim.** The ARDL transformation computes an implied equilibrium
   relationship. It does not by itself prove cointegration or establish a causal
   effect.

The replication is a strong implementation benchmark because it checks a
different software workflow against published numerical results. It does not
replace application-specific identification, stability, specification, and data
quality analysis.

## Reproducibility information

```{r session-info}
sessionInfo()
```

## References

::: {#refs}
:::
